---
title: "An introduction to economic geography"
subtitle: "EK Ricardian models and hat algebra"
author: Ivan Rudik
date: Cornell Dyson
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: GitHub
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' 

---
exclude: true
```{r setup}
options(htmltools.dir.version = FALSE)
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
  }
knitr::opts_chunk$set(echo = TRUE, fig.align="", message = FALSE, warning = FALSE)
  options
})
```
```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "panelset"))

```

---

# Roadmap

1. The economics of space
2. A simple Eaton Kortum Model
3. Exact hat algebra to solve counterfactuals

---

# The economics of space

Capturing space is hard for similar reasons as to why capturing time is hard: lots of states / interactions

--

Historically, very little serious treatment of space

--

.hi-blue["Spatial Impossibility Theorems"] tell us that there's no competitive equilibrium outcome with transport of goods and positive transport costs if space is homogeneous (utilities and production technologies are the same)
  - Factor mobility perfectly substitutes for trade, never incur the trade cost
  

---

# Spatial models

With spatial models we are now going to allow for:

- heterogeneity across space
- trade + frictions
- migration + frictions
- agglomeration economies
- dispersion forces (congestion)

Basically, more consideration of general equilibrium effects

---

# Spatial models

These new approaches generally are considered a part of .hi[economic geography:] the explicit consideration of space

--

Exogenous location-specific characteristics matter

--

Endogenous agglomeration and dispersion forces matter

--

Increasingly important in trade and urban, lots of conceptual/methodological parallels to IO but with a GE twist

--

Often combines granular data (census blocks, households, cell-level) with a theoretical equilibrium model to simulate counterfactuals

---

# Spatial models: macro-ish

These features allow us to better understand the impacts of things like:

- the costs of sea level rise (Desmet et al. 2021)
- climate change and the food problem (Nath 2021)
- climate and agricultural adaptation (Costinot, Donaldson, Smith 2016)
- the geography of environmental regulation (Hollingsworth, Jaworski, Kitchens, and Rudik, 2021)
- the climate adaptation value of migration (Cruz and Rossi-Hansberg 2021)

---

# Spatial models: micro

These features better allow us to understand the impacts of things like:

- the social value of public transit (Severen 2021)
- amenity endogeneity (Almagro and Dominguez-Ilno, 2021)
- the value of density in Berlin (Ahlfeldt et al. 2015)
- the impact of Amazon HQ2 (Dingel and Tintelnot 2021)

---

# Simple set up: a model of market access

We're going to start off with a simple .hi[market access] / .hi-blue[gravity] model

--

The model and code are derived from Donaldson and Hornbeck (2016)

--

It is a version of the widely-used Eaton and Kortum (2002) Ricardian model

<!-- -- -->

<!-- Standard trade models are all isomorphic: same aggregate outcomes despite different microfoundations  -->
<!--   - EK -->
<!--   - Melitz -->
<!--   - Armington -->
<!--   - Krugman -->

---

# Simple set up: a model of market access

Here's what we are going to do:

1. Develop the static model
2. Show how to use .blue[exact hat algebra] to simulate counterfactuals without data on time-invariant exogenous variables (Dekle et al. 2008)
  - Tomorrow's paper uses .blue[dynamic hat algebra] (Caliendo et al. 2019)

--

.hi[New computational piece here:] we will be solving for things up to a .blue[normalization], we won't necessarily be able to get the levels of all variables

---

# Households

$L_i$ households are in location $i = 1,\dots,N$

--

Households supply labor inelastically, and are perfectly mobile across locations (easy to relax)

--

Household utility is
$$U(C_i) = C_i$$

---

# Households

Household utility is
$$U(C_i) = C_i$$

--

Each industry produces a continuum of varieties $\omega \in [0,N]$ (important for the production side) such that $C_i$ is a .pink[constant elasticity of substitution] aggregate of the varieties with elasticity $\sigma$

--

With the CES assumption, we have the .pink[price index] in location $i$ is:
$$P_i = \left(\int_0^N P(\omega)^{1-\sigma}d\omega \right)^{\frac{1}{1-\sigma}}$$


---

# Households: equilibrium utility and wages

The consumer's indirect utility is given by their .pink[real wage:]
$$V_{i} = \frac{w_{i}}{P_{i}}$$

.hi[In equilibrium what must be true about the spatial distribution of V?] 

--

Indirect utility is equalized across locations $\dots$ Why?

--

With perfectly mobile labor, indirect utility is equalized in equilibrium and wages $w_i$ satisfy:
$$w_i = \bar{U} P_i$$

---

# Firms

Firms maximize profits and take prices as given

--

For each location, $z_{i}(\omega)$ is .hi[efficiency]

--

This is the quantity of variety $\omega$ that can be produced using one unit of labor and capital in location $i$

---

# Firms: EK assumption

We assume that $z_{i}(\omega)$ is a random variable distributed according to the Frechet distribution (Eaton and Kortum, 2002):
$$F_{i}\left(z\right) = \exp\left[-T_{i} \left(z\right)^{-\theta}\right]$$


--

This distributional assumption is what will buy us tractability in our model later

---

# Firms: EK assumption

$$F_{i}\left(z\right) = \exp\left[-T_{i} \left(z\right)^{-\theta}\right]$$

--

$\theta > 1$ is the trade elasticity parameter common across all locations and measures the level of intra-industry heterogeneity

--

Smaller $\theta$ $\rightarrow$ more dispersion $\rightarrow$ more heterogeneity $\rightarrow$ stronger .blue[comparative advantage] effects

--

$T_{i}$ measures fundamental productivity

--

Larger values of $T_{i}$ shifts the distribution to the right and corresponds to larger .blue[absolute advantage]

---

# Firms: Production technology

Perfectly competitive firms use labor, capital, and .pink[Cobb-Douglas] technologies to produce different varieties of goods

--

Total output in location $i$ is given by $Y_i = z_i(\omega) L_i^\gamma K_i^{1-\gamma}$

--

$\gamma$ is the labor share and $r_i$ is the rental rate of capital

--

Capital is perfectly mobile so $r_i = r$ for all $i$

--

The input cost of producing variety $\omega$ in location $i$ is:
$$c_i = {(w_i)}^\gamma {(r)}^{1-\gamma}$$

---

# Firms: Trade and prices

Trade costs are .pink[iceberg]

--

Delivering 1 unit of the good from $j$ to $i$ requires shipping $\tau_{ij} \geq 1$ units where $\tau_{ii} = 1$, this represents the costs of shipping/trade


--

The price of a good from $j$ in $i$ is then:
$$P_{ij}(\omega) = \frac{1}{z_i(\omega)}c_i \tau_{ij} = \frac{1}{z_i(\omega)}{(w_i)}^\gamma {(r)}^{1-\gamma} \tau_{ij}$$

--

Buyers purchase from the lowest-cost provider so the price in $i$ is:
$$P_i(\omega) = \min_{j}\left\{P_{ij}(\omega), j=1,\dots,N \right\}$$

---

# Firms: Trade and prices

$P_{ij}(\omega)$ and the Frechet assumption on $z$ give us a distribution of prices faced by $i$ from $j$ sellers $G_{ij}(p)$:
$$Pr[P_{ij} \leq p] \equiv G_{ij}(p) = 1 - F_{j}\left(\frac{1}{p}{(w_i)}^\gamma {(r)}^{1-\gamma} \tau_{ij}\right)$$

--

The lowest price for a variety in $i$ takes on the minimum across all countries $j$ giving us the distribution for prices in $i$ of varieties that $i$ actually buys $G_n(p)$:
$$Pr[P_i \leq p] \equiv G_i(p) = 1 - \prod_{j=1}^N [1- G_{ij}(p)]$$

---

# Firms: Trade and prices

We then get our final expression for the distribution of prices in $i$:
$$G_i(p) = 1 - \exp(-\Phi_i p^\theta) \quad \text{where} \quad \Phi_i = \sum_{j=1}^N T_j (c_j \tau_{ij})^{-\theta}$$

--

$\Phi_i$ summarizes how prices in $i$ are governed by:
  - Fundamental productivity in all $j$
  - Input costs in all $j$
  - Trade barriers with all $j$
  
--
  
What does the distribution look like in autarky or with frictionless trade?

---

# Firms: Prices

Two key properties of $G_i(p)$:

1. The probability that $j$ is the lowest-cost seller in $i$ is:
$$\pi_{ij} = \frac{T_j (c_j \tau_{ij})^{-\theta}}{\sum_{h=1}^N T_h (c_h \tau_{ih})^{-\theta}} \qquad \approx \text{ logit shares}$$
--

<!-- 2. Prices are independent of sources: sellers adjust on the extensive margin by selling more or fewer goods -->
<!-- -- -->

2. The ideal price index in $i$ is $(\sigma < 1+ \theta)$:
$$P_i = \gamma \Phi_i^{-1/\theta}$$


---

# Firms: Prices

With a bit of math we can solve for the .pink[ideal price index] as a function of input costs and parameters:
$$P_{i} = \kappa \left(\sum_{n=1}^N T_{n} \left[{(w_n)}^\gamma \tau_{in} \right]^{-\theta} \right)^{-1/\theta}$$

which is the cost of 1 unit of utility in location $i$, $\kappa$ is a constant

---

# Firms: Trade

$$P_{i} = \kappa \left(\sum_{n=1}^N T_{n} \left[{(w_n)}^\gamma \tau_{in} \right]^{-\theta} \right)^{-1/\theta}$$

Notice $P_i$ is increasing in sellers' trade costs $\tau_{in}$ and input costs $w_n$, and decreasing in productivity $T_n$

--

${(P_{i})}^{-\theta}$ is often called consumer market access $(CMA_i)$ and captures location $i$'s access to cheaper products


---

# Firms: Trade

Next, define a new variable, .pink[expenditure shares:] the share of $i$s total expenditures on products from location $j$: $\lambda_{ij}$

--

Expenditure shares $\equiv$ probability of being the lowest-cost seller:
$$\lambda_{ij} = \frac{T_{j} \left( {(w_j)}^\gamma \tau_{ij}\right)^{-\theta}}{\sum_{h=1}^N T_{h} \left( {(w_h)}^\gamma \tau_{ih}\right)^{-\theta}}$$


---

# Firms: Trade

.hi[Bilateral trade expenditures] of location $i$ on goods from location $j$ are $X_{ij}$ where:

$$\lambda_{ij} = \frac{X_{ij}}{\sum_{h=1}^N X_{ih}} = \frac{X_{ij}}{X_{i}}$$

--

and we can show that:

$$X_{ij} = \kappa T_{j} Y_i\left(\frac{ {(w_j)}^\gamma \tau_{ij}}{{P_{i}}}\right)^{-\theta}$$

$i$ purchases more products from $j$ if the costs of products from $j$ are lower, $i$ is richer, $j$ is more productive


---

# Exact hat algebra

How do we solve this model?

Suppose we want to know the effect of a change in productivity $\hat{T_j} = \frac{T^c_j}{T_j}$

--

Hats denote changes in a variable from factual to counterfactual

--

It turns out it's relatively easy to solve these models in terms of .hi[changes]

--

Representing the model in changes is nice because it will difference out time-invariant variables

--

In this simple model trade costs are the time-invariant variable

---

# Exact hat algebra

First, labor income $w_i L_i = \gamma Y_i$ can be written as a function of the sum of expenditures by $j$ on products from $i$
$$w_i L_i = \gamma \sum_{j=1}^N \lambda_{ji} w_j L_j$$

--

RHS is spending by all locations $j$ on $i$s products

---

# Exact hat algebra

$$\lambda_{ij} = \frac{T_{j} \left( {(w_j)}^\gamma \tau_{ij}\right)^{-\theta}}{\sum_{h=1}^N T_{h} \left( {(w_h)}^\gamma \tau_{ih}\right)^{-\theta}}$$

$$w_i L_i = \gamma \sum_{j=1}^N \lambda_{ji} w_j L_j$$

Suppose we have a productivity shock $\hat{T}_j$, and that $\hat{\tau}_{ij} = \hat{L} = 1$

--

What we are going to do is manipulate these two expressions into .pink[hat form]

---

# Exact hat algebra

Let's start with the counterfactual labor market clearing condition:
$$w^c_i L^c_i = \gamma \sum_{j=1}^N \lambda^c_{ji} w^c_j L^c_j$$

--

$\lambda^c_{ji} w^c_j L^c_j$ is just how much $j$ spends on products from $i$ so that:
$$w^c_i L^c_i = \gamma \sum_{j=1}^N X^c_{ji}$$

---

# Exact hat algebra

Next let's put labor income into hat form:

--

$$\hat{w}_i \hat{L}_i = \frac{w^c_i L^c_i}{w_i L_i} = \gamma \sum_{j=1}^N \frac{X^c_{ji}}{w_i L_i} = \gamma \sum_{j=1}^N \frac{X^c_{ji}}{w_i L_i}\frac{X_{ji}}{X_{ji}} = \gamma \sum_{j=1}^N \frac{X_{ji}}{w_i L_i} \hat{X}_{ji}$$

The change in income due to $\hat{T}_j$ is given by the change in bilateral expenditures, as well as the .pink[factual/observed] income and bilateral expenditures which are just data


---

# Exact hat algebra

Now let's look at the counterfactual expenditure shares:
$$\lambda^c_{ij} = \frac{T^c_{j} \left( {(w^c_j)}^\gamma \tau_{ij}\right)^{-\theta}}{\sum_{h=1}^N T^c_{h} \left( {(w^c_h)}^\gamma \tau_{ih}\right)^{-\theta}}$$

And remember that trade costs .hi[do not] change in this example

--

Next, put it in hat form

---

# Exact hat algebra

$$\hat{\lambda}_{ij} = \frac{\lambda^c_{ij}}{\lambda_{ij}} = \frac{\frac{T^c_{j} \left( {(w^c_j)}^\gamma \tau_{ij}\right)^{-\theta}}{T_{j} \left( {(w_j)}^\gamma \tau_{ij}\right)^{-\theta}}}{\frac{\sum_{h=1}^N T^c_{h} \left( {(w^c_h)}^\gamma \tau_{ih}\right)^{-\theta}}{\sum_{h=1}^N T_{h} \left( {(w_h)}^\gamma \tau_{ih}\right)^{-\theta}}} = \frac{\hat{T}_{j} \left( {(\hat{w}_j)}^\gamma\right)^{-\theta}}{\frac{\sum_{h=1}^N T^c_{h} \left( {(w^c_h)}^\gamma \tau_{ih}\right)^{-\theta}}{\sum_{h=1}^N T_{h} \left( {(w_h)}^\gamma \tau_{ih}\right)^{-\theta}}} = \frac{\hat{T}_{j} \left( {(\hat{w}_j)}^\gamma\right)^{-\theta}}{\hat{P_i}^{-\theta}}$$
The change in expenditure shares is given by the changes in productivity, wages, and prices

--

We still need to solve for $\hat{P}^{-\theta}_i$


---

# Exact hat algebra

We can do so with the price index expression:
$$P_{i} = \kappa \left(\sum_{j=1}^N T_{j} \left[{(w_j)}^\gamma \tau_{ij} \right]^{-\theta} \right)^{-1/\theta}$$

Which gives us:
$$\hat{P}_{i}^{-\theta} = \frac{\left(\sum_{j=1}^N T^c_{j} \left[{(w^c_j)}^\gamma \tau_{ij} \right]^{-\theta} \right)}{\left(\sum_{h=1}^N T_{h} \left[{(w_h)}^\gamma \tau_{ih} \right]^{-\theta} \right)}$$
---

# Exact hat algebra

Next we make a few (potentially odd) substitutions

--

Since $j$ is not in the denominator, we can treat the denominator as a constant and move it inside the sum over $j$:
$$\hat{P}_{i}^{-\theta} = \sum_{j=1}^N \frac{\left(T^c_{j} \left[{(w^c_j)}^\gamma \tau_{ij} \right]^{-\theta} \right)}{\left(\sum_{h=1}^N T_{h} \left[{(w_h)}^\gamma \tau_{ih} \right]^{-\theta} \right)}$$


---

# Exact hat algebra

Next., multiply and divide each term in the sum by the .pink[factual] productivity and wage terms in order to start working toward hat expressions:

$$\hat{P}_{i}^{-\theta} = \sum_{j=1}^N \left[\frac{\left(T^c_{j} \left[{(w^c_j)}^\gamma \tau_{ij} \right]^{-\theta} \right)}{\left(\sum_{h=1}^N T_{h} \left[{(w_h)}^\gamma \tau_{ih} \right]^{-\theta} \right)} \frac{T_j \left[w_j^\gamma\right]^{-\theta}}{T_j \left[w^\gamma_j\right]^{-\theta}} \right]$$

---

# Exact hat algebra

We can then switch out the factual and counterfactual terms up top:

$$\hat{P}_{i}^{-\theta} = \sum_{j=1}^N \left[\frac{\left(T_{j} \left[{(w_j)}^\gamma \tau_{ij} \right]^{-\theta} \right)}{\left(\sum_{h=1}^N T_{h} \left[{(w_h)}^\gamma \tau_{ih} \right]^{-\theta} \right)} \frac{T^c_j \left[{(w^c_j})^\gamma\right]^{-\theta}}{T_j \left[w^\gamma_j\right]^{-\theta}} \right]$$

and this gives us:

$$\hat{P}_{i}^{-\theta} = \sum_{j=1}^N \left[\lambda_{ij} \hat{T}_j \left(\hat{w}_j^\gamma\right)^{-\theta} \right]$$

The change in prices is just data and other hat variables

---

# Exact hat algebra

Here's the three sets of equations again, they are all in terms of changes or data, so we do not need to solve for model fundamentals:

$$\hat{w}_i \hat{L}_i = \gamma \sum_{j=1}^N \frac{X_{ji}}{w_i L_i} \hat{X}_{ji}$$
$$\hat{\lambda}_{ij} = \frac{\hat{T}_{j} \left( {(\hat{w}_j)}^\gamma\right)^{-\theta}}{\hat{P_i}^{-\theta}}$$
$$\hat{P}_{i}^{-\theta} = \sum_{j=1}^N \left[\lambda_{ij} \hat{T}_j \left(\hat{w}_j^\gamma\right)^{-\theta} \right]$$

---

# Exact hat algebra

Given $\hat{T}_i$ we need to identify $\hat{w}_i, \hat{L}_i, \hat{P}_{i}, \hat{\lambda}_{ij}$

--

This is $3N+N^2$ variables, the previous set of 3 equations is only $2N+N^2$

--

Need another $N$ equations to identify the change in equilibrium

--

These $N$ equations are:

$$\frac{\hat{w_i}}{\hat{P}_i} = \frac{\hat{w_j}}{\hat{P}_j} = \widehat{\bar{U}} \,\,\, \forall i,j$$

and

$$\hat{L} = 1$$

---

# Exact hat algebra

$$\hat{P}_{i}^{-\theta} = \sum_{j=1}^N \left[\lambda_{ij} \hat{T}_j \left(\hat{w}_j^\gamma\right)^{-\theta} \right]$$

along with the change in equilibrium requirement that:
$$\frac{\hat{w_i}}{\hat{P}_i} = \widehat{\bar{U}} \,\,\, \forall i$$

allows us to iterate and solve for $\hat{w}_i$ up to a normalization 
---

# Exact hat algebra

$$\hat{\lambda}_{ij} = \frac{\hat{T}_{j} \left( {(\hat{w}_j)}^\gamma\right)^{-\theta}}{\hat{P_i}^{-\theta}}$$

then lets us solve for $\hat{\lambda}_{ij}$ given $\hat{P}_i$ and $\hat{w}_j$

--

Last we need expenditures and labor using the first condition:
$$\hat{w}_i \hat{L}_i = \gamma \sum_{j=1}^N \frac{X_{ji}}{w_i L_i} \hat{X}_{ji}$$

---

# Exact hat algebra

$$\hat{w}_i \hat{L}_i = \gamma  \sum_{j=1}^N \frac{X_{ji}}{w_i L_i} \hat{X}_{ji}$$

Expenditures by $i$ $X_i$ will be equal to labor income so we have that:
$$\hat{X}_i = \gamma  \sum_{j=1}^N \frac{X_{ji}}{w_i L_i} \hat{X}_{ji}$$

--

The definition of expenditure shares gives us:
$$\hat{\lambda}_{ij} = \frac{\hat{X}_{ij}}{\hat{X}_i}$$
---

# Exact hat algebra

Combine these two to get:

$$\frac{\hat{X}_{ih}}{\hat{\lambda}_{ih}} = \gamma  \sum_{j=1}^N \frac{X_{ji}}{w_i L_i} \hat{X}_{ji}$$
which is a set of $n$ equations and $n$ unknown $\hat{X}_{ji}$ for any $h$ we choose

--

We have solved for all the unobserved hat terms given some change in productivity $\hat{T}_i$

--

With these we can simulate any change in productivity, and the only two model parameters we needed to take a stand on were $\gamma$ and $\theta$