---
title: "An introduction to economic geography"
subtitle: "EK Ricardian models and hat algebra"
author: Ivan Rudik
date: Cornell Dyson
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: GitHub
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' 

---
exclude: true
```{r setup}
options(htmltools.dir.version = FALSE)
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
  }
knitr::opts_chunk$set(echo = TRUE, fig.align="", message = FALSE, warning = FALSE)
  options
})
```
```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "panelset"))

```

---

# Roadmap

1. https://arkolakis.com/wp-content/uploads/Economics_and_Space_Unified_at_Last_DEGIT.pdf
1. The economics of space
2. Spatial model simple set up
3. Function iteration solutions
4. Exact hat algebra

---

# The economics of space

Capturing space is hard for similar reasons as to why capturing time is hard: lots of states / interactions

--

Historically, very little serious treatment of space

--

.hi-blue["Spatial Impossibility Theorems"] tell us that there's no competitive equilibrium outcome with transport of goods and positive transport costs if space is homogeneous (utilities and production technologies are the same)
  - Factor mobility perfectly substitutes for trade, never incur the trade cost
  

---

# Spatial models

Spatial models we are now going to allow for:

- heterogeneity across space
- migration
- trade
- agglomeration economies
- congestion forces
- etc

Basically, more consideration of general equilibrium effects

---

# Spatial models

These new approaches generally are considered a part of .hi[economic geography:] the explicit consideration of space

--

Distance, location, etc all matter, but are often ignored

--

Increasingly important in trade and urban, lots of conceptual/methodological parallels to IO but with a GE twist

--

Often combines granular data (census blocks, households, cell-level) with a theoretical equilibrium model to discipline estimation and simulate counterfactuals

---

# Spatial models: macro-ish

These features better allow us to understand the impacts of things like:

- the costs of sea level rise (Desmet et al. 2021)
- climate change and the food problem (Nath 2021)
- climate and agricultural adaptation (Costinot, Donaldson, Smith 2016)
- the geography of environmental regulation (Hollingsworth, Jaworski, Kitchens, and Rudik, 2021)
- the climate adaptation value of migration (Cruz and Rossi-Hansberg 2021)

---

# Spatial models: micro

These features better allow us to understand the impacts of things like:

- the social value of public transit (Severen 2021)
- amenity endogeneity (Almagro and Dominguez-Ilno, 2021)
- the value of density in Berlin (Ahlfeldt et al. 2015)
- the impact of Amazon HQ2 (Dingel and Tintelnot 2021) [.hi[micro-geography]]

---

# Simple set up: a model of market access

We're going to start off with a simple .hi[market access] / .hi-blue[gravity] model

--

The model and code are derived from Donaldson and Hornbeck (2016)

--

It is a version of the canonical Eaton and Kortum (2002) Ricardian model

--

Standard trade models are all isomorphic: same aggregate outcomes despite different microfoundations (Kucheryavyy et al. 2021) 
  - EK
  - Melitz
  - Armington
  - Krugman

---

# Simple set up: a model of market access

Here's what we are going to do:

1. Develop the static model
2. Show how to solve for model unknowns using data and calibration of parameters 
3. Show how to use .blue[exact hat algebra] to simulate counterfactuals without needing to observe time-invariant exogenous variables (Dekle et al. 2008)
  - Empirical seminar talk uses .blue[dynamic hat algebra] (Caliendo et al. 2019)

--

.hi[New computational piece here:] we will be solving for things up to a .blue[normalization], we won't necessarily be able to get the levels of all variables

---

# Households

$L_i$ households are in location $i = 1,\dots,N$

--

Households supply labor inelastically, and are perfectly mobile across locations

--

Household utility is
$$U(C_i) = C_i$$

---

# Households

Household utility is
$$U(C_i) = C_i$$

--

Each industry produces a continuum of varieties $\omega \in [0,N]$ (important for the production side) such that $C_i$ is a .pink[constant elasticity of substitution] aggregate of the varieties with elasticity $\sigma$

--

With the CES assumption, we have the .pink[price index] in location $i$ is:
$$P_i = \left(\int_0^N P(\omega)^{1-\sigma}d\omega \right)^{\frac{1}{1-\sigma}}$$


---

# Households: equilibrium utility and wages

With Cobb-Douglas preferences, a consumer's indirect utility is given by their .pink[real wage:]
$$V_{i} = \frac{w_{i}}{P_{i}}$$

.hi[In equilibrium what must be true?] 

--

Indirect utility is equalized across locations $\dots$ Why?

--

With perfectly mobile labor, indirect utility is equalized in equilibrium and wages $w_i$ satisfy:
$$w_i = \bar{U} P_i$$

---

# Firms

Firms maximize profits and take prices as given

--

For each location, $z_{i}(\omega)$ is .hi[efficiency]

--

This is the quantity of variety $\omega$ that can be produced using one unit of labor in location $i$

---

# Firms: EK assumption

We assume that $z_{i}(\omega)$ is a random variable distributed according to the Frechet distribution (Eaton and Kortum, 2002):
$$F_{i}\left(z\right) = \exp\left[-T_{i} \left(z\right)^{-\theta}\right]$$


--

This distributional assumption is what will buy us tractability in our model later

---

# Firms: EK assumption

$$F_{i}\left(z\right) = \exp\left[-T_{i} \left(z\right)^{-\theta}\right]$$

--

$\theta > 1$ is the trade elasticity parameter common across all states and counties and measures the level of intra-industry heterogeneity

--

Smaller $\theta$ $\rightarrow$ more dispersion $\rightarrow$ more heterogeneity $\rightarrow$ stronger comparative advantage effects

--

$T_{i}$ measures fundamental productivity

--

Larger values of $T_{i}$ increase the probability of larger efficiency draws $z_{i}(\omega)$ and indicate larger absolute advantage

---

# Firms: Production technology

Perfectly competitive firms use labor, capital, and .pink[Cobb-Douglas] technologies to produce different varieties of goods

--

Total output in location $i$ is given by $Y_i = z_i(\omega) L_i^\gamma K_i^{1-\gamma}$

--

$\gamma_i$ is the share of value added and $r_i$ is the rental rate of capital

--

Capital is perfectly mobile so $r_i = r$ for all $i$

--

The input cost of producing variety $\omega$ in location $i$ is:
$$c_i = {(w_i)}^\gamma {(r)}^{1-\gamma}$$

---

# Firms: Trade and prices

Trade costs are .pink[iceberg]

--

Delivering 1 unit of the good from $j$ to $i$ requires shipping $\tau_{ij} \geq 1$ units, this represents the costs of shipping/trade


--

The price of a good from $j$ in $i$ is then:
$$P_{ij}(\omega) = \frac{1}{z_i(\omega)}c_i(\omega) \tau_{ij} = \frac{1}{z_i(\omega)}{(w_i)}^\gamma {(r)}^{1-\gamma} \tau_{ij}$$

--

Buyers purchase from the lowest-cost provider so the price in $i$ is:
$$P_i(\omega) = \min_{j}\left\{P_{ij}(\omega), j=1,\dots,N \right\}$$

---

# Firms: Trade and prices

$P_{ij}(\omega)$ and the Frechet assumption on $z$ give us a distribution of prices faced by $i$ from $j$ sellers $G_{ij}(p)$:
$$Pr[P_{ij} \leq p] \equiv G_{ij}(p) = 1 - F_{j}\left(\frac{1}{p}{(w_i)}^\gamma {(r)}^{1-\gamma} \tau_{ij}\right)$$

--

The lowest price for a variety in $i$ takes on the minimum across all countries $j$ giving us the distribution for prices in $i$ of varieties that $i$ actually buys $G_n(p)$:
$$Pr[P_i \leq p] \equiv G_i(p) = 1 - \prod_{j=1}^N [1- G_{ij}(p)]$$

---

# Firms: Trade and prices

We then get our final expression for the distribution of prices in $i$:
$$G_i(p) = 1 - \exp(-\Phi_i p^\theta) \quad \text{where} \quad \Phi_i = \sum_{j=1}^N T_j (c_j \tau_{ij})^{-\theta}$$

$\Phi_i$ summarizes how:
  - Technologies in all $j$
  - Input costs in all $j$
  - Trade barriers with all $j$
  
Govern prices in $i$

---

# Firms: Prices

Three key properties of $G_i(p)$:

1. The probability that $j$ is the lowest-cost seller in $i$ is:
$$\pi_{ij} = \frac{T_j (c_j \tau_{ij})^{-\theta}}{\sum_{j=1}^N T_j (c_j \tau_{ij})^{-\theta}} \qquad \approx \text{ logit shares}$$

2. Prices are independent of sources: sellers adjust on the extensive margin by selling more or fewer goods

3. The ideal price index in $i$ is $(\sigma < 1+ \theta)$:
$$P_i = \gamma \Phi_i^{-1/\theta}$$


---

# Firms: Prices

With a bit of math we can solve for the .pink[ideal price index] as a function of input costs and parameters:
$$P_{i} = \kappa \left(\sum_{n=1}^N T_{n} \left[{(w_n)}^\gamma \tau_{in} \right]^{-\theta} \right)^{-1/\theta}$$

which is the cost of 1 unit of utility in location $i$, $\kappa$ is a constant

---

# Firms: Trade

$$P_{i} = \kappa \left(\sum_{n=1}^N T_{n} \left[{(w_n)}^\gamma \tau_{in} \right]^{-\theta} \right)^{-1/\theta}$$

Notice $P_i$ is increasing in sellers' trade costs $\tau_{in}$ and input costs $w_n$, and decreasing in productivity $T_n$

--

${(P_{i})}^{-\theta}$ is often called consumer market access $(CMA_i)$ and captures location $i$'s access to cheaper products

---

# Firms: Trade

.hi[Bilateral trade expenditures] of location $i$ on goods from location $j$ are:
$$X_{ij} = \kappa T_{j} Y_i\left(\frac{ {(w_j)}^\gamma \tau_{ij}}{{P_{i}}}\right)^{-\theta}$$


Prop 2 $\rightarrow$ average expenditure per good $\perp$ source, $\rightarrow$ Prop 1 $\rightarrow$ fraction of quantities $\leftrightarrow$ fraction of expenditures

--

$i$ purchases more products from $j$ if:

- the costs of products from $j$ are lower, $i$ is richer $Y_i$, consumer market access ${(P_{i})}^{-\theta}$ is lower


---

# Firms: Market access

Firms also have a .hi[market access] term $FMA_j$ that we can derive from combining the equilibrium real wage, price index, and trade expenditures expressions:
$$FMA_j = \sum_{i=1}^N \frac{\tau_{ij}^{-\theta}Y_i}{CMA_i}$$

--

Firm market access captures the firm side: their access to higher price markets

---

# Firms: Market access

$$FMA_j = \sum_{i=1}^N \frac{\tau_{ij}^{-\theta}Y_i}{CMA_i}$$

FMA is higher if:

- Trade costs are lower $(\tau_{ij})$
- Other locations are richer $(Y_i)$


---

# We did it

So that is the model

--

Now we need to solve it

--

One of the key pieces in our model are the .hi[market access terms]

--

It turns out, FMA and CMA are .hi-blue[equivalent up to a normalization by a constant:]
$$MA_i \equiv FMA_i = \rho CMA_i$$


---

# Solving the model

Typically we solve for the model equilibrium by using equilibrium conditions to solve for endogenous quantities

--

Here we don't observe all exogenous quantities

--

We can calibrate/estimate trade costs, labor share

--

We still need the .pink[the N unobserved and exogenous] $T_i$ .pink[terms]

---

# Solving the model

How do we solve for the $T_i$ terms?

--

We observe a number of endogenous variables (labor, wages) of length $>$ N

--

If we assume the real world is in a static equilibrium we can use equilibrium conditions to solve for $T_i$

--

Before we are able to solve for $T_i$ we need to solve for one more endogenous variable: market access / price index

---

# Solving the model: market access

$$FMA_j = \sum_{i=1}^N \frac{\tau_{ij}^{-\theta}Y_i}{CMA_i}$$

--

using the equivalency between FMA and CMA, and that $Y_i = \frac{w_i L_i}{\gamma}$ from labor market clearing we have our condition allowing us to solve for market access:
$$MA_j = \rho \sum_{i=1}^N \tau_{ij}^{-\theta} MA_i^{\frac{-(1+\theta)}{\theta}} L_i$$

For $j=1,\dots,N$ this is a system of N equations

---

# Solving the model: market access

$$MA_j = \rho \sum_{i=1}^N \tau_{ij}^{-\theta} MA_i^{\frac{-(1+\theta)}{\theta}} L_i$$

$N$ equations, $N$ unknown MA terms, everything else is a parameter we can estimate $(\tau_{ij}, \theta)$, data $L_i$, or a normalizing constant $\rho$

--

.hi[Fujimoto-Krause] shows that we can solve for the vector $MA$ up to this normalization using function iteration, but where our update is:
$$MA_{new} = \frac{RHS(MA_{old})}{||RHS(MA_{old})||}$$


---

# Fujimoto-Krause in matrix form

```
function solve_price(data, params, labor)

    # set up intial guesses
    ma_temp = collect(1:size(data.L)[1])
    ma_new = data.tau_theta * (ma_temp .^ (-(1 + params.theta) / params.theta) .* labor)
    ma_temp = ma_new / sqrt(sum(ma_new .^ 2))
    error = Inf

    # use Fujimoto-Krause to solve for MA up to a normalization
    while error > params.tol * params.damping
        ma_new = data.tau_theta * (ma_temp .^ (-(1 + params.theta) / params.theta) .* labor) # RHS of MA equation
        ma_new = ma_new / sqrt(sum(ma_new .^ 2)) # normalize
        error = sqrt.(sum((ma_new .- ma_temp) .^ 2))
        ma_temp = params.damping * ma_new + (1 - params.damping) * ma_temp # damp the update
    end
    
    # Convert to price index
    price = ma_new .^ (-1 / params.theta)
    return price
    
end
```

---

# Solving the model: productivity

Now we need to solve for productivity $T_j$

--

How do we do this?

--

Recall bilateral expenditures are:
$$X_{ij} = \kappa T_{j} Y_i\left(\frac{ {(w_j)}^\gamma \tau_{ij}}{{P_{i}}}\right)^{-\theta}$$

and the price index is:

$$MA_i = P_{i}^{-\theta} = \kappa \left(\sum_{n=1}^N T_{n} \left[{(w_n)}^\gamma \tau_{in} \right]^{-\theta} \right)$$
 
---

# Solving the model: productivity

$$X_{ij} = \kappa T_{j} Y_i\left(\frac{ {(w_j)}^\gamma \tau_{ij}}{{P_{i}}}\right)^{-\theta}$$


Plug in $P_i$, and use $\bar{U} = \frac{w_i}{P_i}$ to get an expression for equilibrium labor allocations as a function of productivity and market access:
$$L_j = \underbrace{\kappa \gamma T_j \bar{U}^{-(1+\gamma \theta)}}_{\eta_j} MA_j^{\frac{\theta(1+\gamma)+1}{\theta}}$$
where $\kappa$ is some constant, and $\eta_j$ is $T_j$ up to a normalization by a constant


---

# Solving the model: productivity


$$L_j = \eta_j MA_j^{\frac{\theta(1+\gamma)+1}{\theta}}$$

--

We can solve for $\eta_j$ which is just $T_j$ up to a normalization by a constant $\kappa \gamma \bar{U}^{-(1+\gamma \theta)}$
$$\eta_j = L_j MA_j^{-\frac{\theta(1+\gamma)+1}{\theta}}$$

--

Labor is higher areas with greater market access (cheaper goods), or higher productivity (higher wage)

---

# Solving the model: productivity

```
function solve_productivity(data, params)

    # Iterate on equation to solve for market access
    # Then invert to recover P
    price = solve_price(data, params, data.L)

    # recover T up to a normalization
    T =
        exp.(
            log.(data.L) .-
            ((1 + params.theta * (1 + params.gamma)) / params.theta) .*
            log.(price .^ (-params.theta)),
        )

    return T

end
```

--

Now we have all the unknown exogenous pieces of the model and can simulate it given changes in model parameters

---

# Simulating the model

Suppose we want to measure the effect of a change in productivity to $T^c_j$ where superscript $c$ denotes counterfactual quantities

--

We have 2N+1 unknowns: $MA^c_j, L^c_j, \bar{U}^c$ for $j=1,\dots,N$

---

# Simulating the model

First rearrange the labor expression to solve for market access:
$$MA^c_j = \left[\frac{L^c_j}{\eta^c_j}\right]^{\frac{\theta}{\theta(1 + \gamma) + 1}}$$

where $\eta^c_j$ is a function of $T^c_j$

--

This is a set of $N$ equations that pins down MA conditional on L

---

# Simulating the model

$$MA^c_j = \left[\frac{L^c_j}{\eta^c_j}\right]^{\frac{\theta}{\theta(1 + \gamma) + 1}}$$

--

Next we can insert the $MA^c_j$ terms into the MA fixed point equation:
$$MA^c_j = \rho \sum_{i=1}^N \tau_{ij}^{-\theta} {MA^c_i}^{\frac{-(1+\theta)}{\theta}} L^c_i$$

--

This is another $N$ equations that now pins down L

--

We still need to pin down $\bar{U}^c$


---

# Simulating the model

Pinning down $\bar{U}^c$ is where the *up to a normalization* part comes in

--

The pinned down $L^c_j$s above won't necessarily sum up to $\sum_{i=1}^N L_i$

--

We pin down $\bar{U}^c$ by inflating all the $L^c_j$ terms by some constant factor $\zeta$ so they sum up to the factual level of labor $\sum_{i=1}^N L_i$

--

We are just imposing the last remaining equation, the labor resource constraint:
$$\sum_{i=1}^N L^c_i = \sum_{i=1}^N L_i$$

---

# Simulating the model

$$MA^c_j = \left[\frac{L^c_j}{\eta^c_j}\right]^{\frac{\theta}{\theta(1 + \gamma) + 1}}$$

$$MA^c_j = \rho \sum_{i=1}^N \tau_{ij}^{-\theta} {MA^c_i}^{\frac{-(1+\theta)}{\theta}} L^c_i$$

$$\sum_{i=1}^N L^c_i = \sum_{i=1}^N L_i$$

2N+1 equations and 2N+1 unknowns, we can solve this system of equations to get our new equilibrium under $\bar{T}^c_j$


---

# Simulating the model

We can then compute the change in welfare from this new equilibrium:
$$\frac{\bar{U}^c}{\bar{U}} = \left({\eta^c / \eta}\right)^{-1/(1+\gamma \theta)}$$

--

All the constant terms that we never explicitly solved for in $\eta$ cancel out so we can still identify the .hi[change] in welfare and employment

--

Now lets look at the code


---

# Simulating the model: solving for a change in welfare

Suppose we change eta from `eta_bar` to `eta_counter`

--

Our code might look like:

```{julia, eval = F}
function solve_counterfactual_base(eta_bar, data, params, regulation_counter)

    eta_counter = eta_bar .* regulation_counter

    # solve for labor allocation
    labor_counter = solve_labor(data, params, eta_counter)
    .
    .
```

What does `solve_labor` do?

---

# Simulating the model: solving for labor

```{julia, eval = F}
function solve_labor(data, params, eta_counter)

    # insert L into market access
    g = 1 + params.theta * (1 + params.gamma)
    L_temp = data.L
    L_new =
        eta_counter .*
        (data.tau_theta * ((L_temp ./ eta_counter) .^ (-(1 + params.theta) / g) .* L_temp)) .^
        (g / params.theta)
    L_temp = L_new / sqrt(sum(L_new .^ 2))
    error = Inf
    .
    .
    # next do function iteration
```

---

# Simulating the model: solving for labor

```{julia, eval = F}
    .
    .
    # use Fujimoto-Krause to solve for L up to a normalization
    while error > params.tol
        L_new =
            eta_counter .*
            (data.tau_theta * ((L_temp ./ eta_counter) .^ (-(1 + params.theta) / g) .* L_temp)) .^
            (g / params.theta)
        L_new = L_new / sqrt(sum(L_new .^ 2))
        error = sqrt.(sum((L_new .- L_temp) .^ 2))
        L_temp = L_new
    end
    return L_new
end
```


---

# Simulating the model: solving for a change in welfare

```{julia, eval = F}
    .
    labor_counter = solve_labor(data, params, T_counter)
    labor_counter = labor_counter ./ sum(labor_counter) .* sum(data.L) # normalize
    price_counter = solve_price(data, params, labor_counter) # get price index

    eta_counter = # with P^c, L^c we can solve for eta^c
        exp.(log.(labor_counter) .-
            ((1 + params.theta * (1 + params.gamma)) / params.theta) .*
            log.(price_counter .^ (-params.theta)),
        ) ./ regulation_counter
    delta_U = (eta_counter ./ eta_bar) .^ (-1 / (1 + params.gamma * params.theta))

    results = DataFrame(
        eta_bar = vec(eta_bar), T_bar_counter = vec(eta_bar_counter),
        cf_reg = vec(regulation_counter),
        labor_before = vec(data.L), labor_after = vec(labor_counter),
        labor_change = vec(labor_counter ./ data.L), welfare_change = vec(delta_U))

    return results
 
end
```

---

# Exact hat algebra

We can solve this in a different way with fewer assumptions

Suppose we want to know the effect of a change in productivity $\hat{T_j} = \frac{T^c_j}{T_j}$

--

Hats denote changes in a variable from factual to counterfactual

--

It turns out it's relatively easy to solve these models in terms of .hi[changes]

--

Representing the model in changes is nice because it will difference out time-invariant variables

---

# Exact hat algebra

First we need to define a new variable, .pink[expenditure shares:]
$$\lambda_{ij} = \frac{X_{ij}}{\sum_{h=1}^N X_{ih}} = \frac{X_{ij}}{X_{i}}$$

--

This is just the share of total expenditures on products from location $j$

--

Substitute in for the $X_{ih}$ terms from the previous expression to get:
$$\lambda_{ij} = \frac{T_{j} \left( {(w_j)}^\gamma \tau_{ij}\right)^{-\theta}}{\sum_{h=1}^N T_{h} \left( {(w_h)}^\gamma \tau_{ih}\right)^{-\theta}}$$

---

# Exact hat algebra

Next, labor income $w_i L_i = \gamma Y_i$ can be written as a function of the sum of expenditures by $j$ on products from $i$
$$w_i L_i = \gamma \sum_{j=1}^N \lambda_{ji} w_j L_j$$

--

RHS is spending by all locations $j$ on $i$s products, scaled by the labor share

---

# Exact hat algebra

$$\lambda_{ij} = \frac{T_{j} \left( {(w_j)}^\gamma \tau_{ij}\right)^{-\theta}}{\sum_{h=1}^N T_{h} \left( {(w_h)}^\gamma \tau_{ih}\right)^{-\theta}}$$

$$w_i L_i = \gamma \sum_{j=1}^N \lambda_{ji} w_j L_j$$

Suppose we have a productivity shock $\hat{T}_j$, and that $\hat{\tau}_{ij} = \hat{L} = 1$

--

What we are going to do is manipulate these two expressions into .pink[hat form]

---

# Exact hat algebra

Let's start with the counterfactual labor market clearing condition:
$$w^c_i L^c_i = \gamma \sum_{j=1}^N \lambda^c_{ji} w^c_j L^c_j$$

$\lambda^c_{ji} w^c_j L^c_j$ is just how much $j$ spends on products from $i$ so that:
$$w^c_i L^c_i = \gamma \sum_{j=1}^N X^c_{ji}$$

---

# Exact hat algebra

Next let's put labor income into hat form:
$$\hat{w}_i \hat{L}_i = \frac{w^c_i L^c_i}{w_i L_i} = \gamma \sum_{j=1}^N \frac{X^c_{ji}}{w_i L_i} = \gamma \sum_{j=1}^N \frac{X^c_{ji}}{w_i L_i}\frac{X_{ji}}{X_{ji}} = \gamma \sum_{j=1}^N \frac{X_{ji}}{w_i L_i} \hat{X}_{ji}$$

The change in income due to $\hat{T}_j$ is given by the change in bilateral expenditures, as well as the .pink[factual/observed] income and bilateral expenditures which are just data


---

# Exact hat algebra

Now let's look at the counterfactual expenditure shares:
$$\lambda^c_{ij} = \frac{T^c_{j} \left( {(w^c_j)}^\gamma \tau_{ij}\right)^{-\theta}}{\sum_{h=1}^N T^c_{h} \left( {(w^c_h)}^\gamma \tau_{ih}\right)^{-\theta}}$$

And remember that trade costs .hi[do not] change in this example

--

Next, put it in hat form

---

# Exact hat algebra

$$\hat{\lambda}_{ij} = \frac{\lambda^c_{ij}}{\lambda_{ij}} = \frac{\frac{T^c_{j} \left( {(w^c_j)}^\gamma \tau_{ij}\right)^{-\theta}}{T_{j} \left( {(w_j)}^\gamma \tau_{ij}\right)^{-\theta}}}{\frac{\sum_{h=1}^N T^c_{h} \left( {(w^c_h)}^\gamma \tau_{ih}\right)^{-\theta}}{\sum_{h=1}^N T_{h} \left( {(w_h)}^\gamma \tau_{ih}\right)^{-\theta}}} = \frac{\hat{T}_{j} \left( {(\hat{w}_j)}^\gamma\right)^{-\theta}}{\frac{\sum_{h=1}^N T^c_{h} \left( {(w^c_h)}^\gamma \tau_{ih}\right)^{-\theta}}{\sum_{h=1}^N T_{h} \left( {(w_h)}^\gamma \tau_{ih}\right)^{-\theta}}} = \frac{\hat{T}_{j} \left( {(\hat{w}_j)}^\gamma\right)^{-\theta}}{\hat{P_i}^{-\theta}}$$
The change in expenditure shares is given by the changes in productivity, wages, and prices

--

We still need to solve for $\hat{P}^{-\theta}_i$


---

# Exact hat algebra

We can do so with the price index expression:
$$P_{i} = \kappa \left(\sum_{j=1}^N T_{j} \left[{(w_j)}^\gamma \tau_{ij} \right]^{-\theta} \right)^{-1/\theta}$$

Which gives us:
$$\hat{P}_{i}^{-\theta} = \frac{\left(\sum_{j=1}^N T^c_{j} \left[{(w^c_j)}^\gamma \tau_{ij} \right]^{-\theta} \right)}{\left(\sum_{h=1}^N T_{h} \left[{(w_h)}^\gamma \tau_{ih} \right]^{-\theta} \right)}$$
---

# Exact hat algebra

Next we make a few (potentially odd) substitutions

--

Since $j$ is not in the denominator, we can treat the denominator as a constant and move it inside the sum over $j$:
$$\hat{P}_{i}^{-\theta} = \sum_{j=1}^N \frac{\left(T^c_{j} \left[{(w^c_j)}^\gamma \tau_{ij} \right]^{-\theta} \right)}{\left(\sum_{h=1}^N T_{h} \left[{(w_h)}^\gamma \tau_{ih} \right]^{-\theta} \right)}$$


---

# Exact hat algebra

Next., multiply and divide each term in the sum by the .pink[factual] productivity and wage terms in order to start working toward hat expressions:

$$\hat{P}_{i}^{-\theta} = \sum_{j=1}^N \left[\frac{\left(T^c_{j} \left[{(w^c_j)}^\gamma \tau_{ij} \right]^{-\theta} \right)}{\left(\sum_{h=1}^N T_{h} \left[{(w_h)}^\gamma \tau_{ih} \right]^{-\theta} \right)} \frac{T_j \left[w_j^\gamma\right]^{-\theta}}{T_j \left[w^\gamma_j\right]^{-\theta}} \right]$$

---

# Exact hat algebra

We can then switch out the factual and counterfactual terms up top:

$$\hat{P}_{i}^{-\theta} = \sum_{j=1}^N \left[\frac{\left(T_{j} \left[{(w_j)}^\gamma \tau_{ij} \right]^{-\theta} \right)}{\left(\sum_{h=1}^N T_{h} \left[{(w_h)}^\gamma \tau_{ih} \right]^{-\theta} \right)} \frac{T^c_j \left[{(w^c_j})^\gamma\right]^{-\theta}}{T_j \left[w^\gamma_j\right]^{-\theta}} \right]$$

and this gives us:

$$\hat{P}_{i}^{-\theta} = \sum_{j=1}^N \left[\lambda_{ij} \hat{T}_j \left(\hat{w}_j^\gamma\right)^{-\theta} \right]$$

The change in prices is just data and other hat variables

---

# Exact hat algebra

Here's the three sets of equations again, they are all in terms of changes or data, so we do not need to solve for model fundamentals:

$$\hat{w}_i \hat{L}_i = \gamma \sum_{j=1}^N \frac{X_{ji}}{w_i L_i} \hat{X}_{ji}$$
$$\hat{\lambda}_{ij} = \frac{\hat{T}_{j} \left( {(\hat{w}_j)}^\gamma\right)^{-\theta}}{\hat{P_i}^{-\theta}}$$
$$\hat{P}_{i}^{-\theta} = \sum_{j=1}^N \left[\lambda_{ij} \hat{T}_j \left(\hat{w}_j^\gamma\right)^{-\theta} \right]$$

---

# Exact hat algebra

Given $\hat{T}_i$ we need to identify $\hat{w}_i, \hat{L}_i, \hat{P}_{i}, \hat{\lambda}_{ij}$

--

This is $3N+N^2$ variables, the previous set of 3 equations is only $2N+N^2$

--

Need another $N$ equations to identify the change in equilibrium

--

These $N$ equations are:

$$\frac{\hat{w_i}}{\hat{P}_i} = \frac{\hat{w_j}}{\hat{P}_j} = \widehat{\bar{U}} \,\,\, \forall i,j$$

and

$$\hat{L} = 1$$

---

# Exact hat algebra

$$\hat{P}_{i}^{-\theta} = \sum_{j=1}^N \left[\lambda_{ij} \hat{T}_j \left(\hat{w}_j^\gamma\right)^{-\theta} \right]$$

along with the change in equilibrium requirement that:
$$\frac{\hat{w_i}}{\hat{P}_i} = \widehat{\bar{U}} \,\,\, \forall i$$

allows us to iterate and solve for $\hat{P}_i$ up to a normalization since $\hat{L} = \frac{\sum_{i=1}^N L^c_i}{\sum_{i=1}^N L_i} = 1$ will later pin down $\widehat{\bar{U}}$

---

# Exact hat algebra

$$\hat{\lambda}_{ij} = \frac{\hat{T}_{j} \left( {(\hat{w}_j)}^\gamma\right)^{-\theta}}{\hat{P_i}^{-\theta}}$$

then lets us solve for $\hat{\lambda}_{ij}$ given $\hat{P}_i$ and $\hat{w}_j$

--

Last we need expenditures and labor using the first condition:
$$\hat{w}_i \hat{L}_i = \gamma \sum_{j=1}^N \frac{X_{ji}}{w_i L_i} \hat{X}_{ji}$$

---

# Exact hat algebra

$$\hat{w}_i \hat{L}_i = \gamma  \sum_{j=1}^N \frac{X_{ji}}{w_i L_i} \hat{X}_{ji}$$

Expenditures by $i$ $X_i$ will be equal to labor income so we have that:
$$\hat{X}_i = \gamma  \sum_{j=1}^N \frac{X_{ji}}{w_i L_i} \hat{X}_{ji}$$

--

The definition of expenditure shares gives us:
$$\hat{\lambda}_{ij} = \frac{\hat{X}_{ij}}{\hat{X}_i}$$
---

# Exact hat algebra

Combine these two to get:

$$\frac{\hat{X}_{ih}}{\hat{\lambda}_{ih}} = \gamma  \sum_{j=1}^N \frac{X_{ji}}{w_i L_i} \hat{X}_{ji}$$
which is a set of $n$ equations and $n$ unknown $\hat{X}_{ji}$ for any $h$ we choose

--

We have solved for all the unobserved hat terms given some change in productivity $\hat{T}_i$

--

With these we can simulate any change in productivity, and the only two model parameters we needed to take a stand on were $\gamma$ and $\theta$